\begin{table}[t]
\centering
\begin{tabular}{c|c|c}
Component & Files & Size (in lines)\\\hline
VST Examples (\S\ref{sec:vstgraphmark},\S\ref{sec:application}) & 13 & 3,290 \\
Integration into Floyd (\S\ref{sec:vstlocalunlocal}) & 12 & 1,917 \\
H/S Example (\S\ref{sec:hipsleek}) & 2 & 443 \\
Modifications to H/S (\S\ref{sec:hipsleek}) & ~~51\tablefootnote{H/S files are modified, not necessarily freshly created.} & 2,500 \\
Math Graph (\S\ref{sec:mathgraph}) & 19 & 12,723\\
Spatial Graph (\$\ref{sec:spacegraph}) & 10 & 6,458 \\
Memory Model \& Logic & 13 & 2,373 \\
Common Utilities & 10 & 2,842 \\\hline\hline
Total Development & 130 & 32,546 \\
\end{tabular}
\caption{Size of our codebase}
\label{tab:codebase}
\end{table}

All of our results in this paper have been machine-checked.  The bulk of our development was checked in Coq, although a 55-line file (shown in Figure~\ref{fig:hipmarkgraph}) was checked in HIP/SLEEK.  Our modifications to H/S itself were not machine-checked since H/S does not have a mechanized soundness proof.
Although the size of a development does not perfectly match with that development's importance or implementation difficulty, we present it nonetheless in Table~\ref{tab:codebase}, organized roughly by the paper section corresponding to each development.  For comparison, verifying a simple list-based merge sort in VST takes 600 lines.  It took fewer than 400 lines of Coq to verify the \li{mark} algorithm in H/S, indicating that a large portion of the codebase is shared between VST and H/S.  For context, VST and H/S have approximately 140k and 179k LOC.

\hide{
\paragraph{Size of Coq mathgraph \S\ref{sec:mathgraph}.}
Graph folder has 18 files in total with 10,919 lines in total

\paragraph{Size of Coq spacegraph \S\ref{sec:spacegraph}.}

not including the connection to H/S or VST

include \li{Graph.v} and \li{GraphBi.v} from \li{msl_application} since they do not depend on the underlying VST model?

What is in \li{ramification_lemmas}?  %Any other parts of VST ``standard'' from our work?

\paragraph{Integration into Floyd (\S\ref{sec:vst}).}: ?
Size of additions to VST logic model: ?

\paragraph{Modifications to HIP/SLEEK.}
H/S code: approximately 2,500 lines of code across 51 files
Size of extra H/S memory model:
\li{alg_seplog_direct.v}  52 lines
\li{overlapping_direct.v} 442 lines  {\color{magenta}How do we handle alignment?}
\li{precise_direct.v} 111 lines
other files?

\paragraph{Size of VST examples~\S\ref{sec:application}.}

Note: graph, graphbi (in space), graphmark, graphbimark are shared in \li{msl_application}

mark graph  19 + 402 + 161 + 246 = 828
mark dag  19 + 402 + 161 + 210
Note: mark dag shares 19 + 402 + 161 with mark graph

copy   459 + 19 + 161 + 388 = 1,027 (need to add files from \li{msl_applicatin})
dispose   475 + 18 + 544 = 1,037 (need to add files from \li{data_structure})
What do copy or dispose share with mark or with each other?

total: 1,038 (marks) + 2,064 (copy/dispose, assuming no duplication) = 3,102 lines.  12 files, assuming no sharing for copy/dispose (with each other or with mark)

\paragraph{Size of H/S example.}
main file: 54 lines (Figure~\ref{fig:hipmarkgraph}), \li{Module Type} generated by H/S (Figure~\ref{fig:hipcoqfile}) is 30 lines, Coq \li{Module} matching this \li{Module Type} is 358 lines.
total: 2 human-generated files, 429 lines

\paragraph{Size of total development.} ?
}
