\paragraph{Comparison with Hobor and Villard \cite{hobor:ramification}.}
Hobor and Villiard verified graph algorithms using the \infrulestyle{Ramify} rule.  We have upgraded this rule to better handle modified program variables and existential quantifiers in postconditions; they avoided these challenges by proposing a variant of \infrulestyle{Ramify} called \infrulestyle{RamifyAssign}, which could reason about the special case of a single assignment $\li{x=}f(\ldots)$, assuming the verifier can make the local program translation to $\li{x'=}f(\ldots)\li{; x=x'}$, where \li{x'} is fresh.  They proposed no way to verify unmodified program code, to modify program variables inside nested localization blocks, or to do a ramification across multiple assignments as in lines~\ref{code:markbeforetripleramify}--\ref{code:markaftertripleramify} of figure~\ref{fig:markgraph}.  Hobor and Villard avoided existentials in localized postconditions by defining all their mathematical operations (\emph{e.g.} $\m{mark}$) as functions rather than as relations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% Edit 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% Edit 1: Qinxiang's proposal starts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\iffalse
Our development is entirely machine-checked~(\S\ref{sec:development}) which revealed some tricky technique details. Hobor and Villard fell into the trap of defining spatial graphs recursively~(\S\ref{sec:fixpointfail}); unfortunately other members of the research community have since followed them in.  We exposed this error and provided a sound, general, and highly modular graph framework that works smoothly in a mechanized context~(\S\ref{sec:mathgraph},\S\ref{sec:spacegraph}).
% and quite general
%definition for \p{graph}~(\S\ref{sec:goodgraph}) that recovers fold/unfold reasoning.
\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% Edit 1: Qinxiang's proposal ends
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% Edit 1: Original version starts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\iftrue
Hobor and Villard treated math graphs very simply, as triples $(V,E,L)$ of vertices, edges, and a labeling function on vertices.  Vertices had no more than two neighbors.  In contrast, our mathematical graph framework~(\S\ref{sec:mathgraph}) is very modular and general and has been tuned to work smoothly in a mechanized context.

Hobor and Villard erroneously defined spatial graphs recursively~(\S\ref{sec:fixpointfail}); unfortunately other members of the research community have since followed them in, \emph{e.g.}~\cite{RaadVG15}.  We exposed this error and provided a sound and quite general definition for \p{graph}~(\S\ref{sec:goodgraph}) that recovers fold/unfold reasoning.  We developed a much more general and more modular set of related lemmas and connect our spatial reasoning to two very different verification tools~(\S\ref{sec:ramifylib}), VST~(\S\ref{sec:vst}) and HIP/SLEEK~(\S\ref{sec:hipsleek}).  Our development is entirely machine-checked~(\S\ref{sec:development}) whereas they used only pen and paper.
\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%% Edit 1: Original version ends
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Other verification of graph algorithms and/or $**$.}
Yang's verification of the Schorr-Waite algorithm is a landmark in the early separation logic literature~\cite{hongseok:phd}.  Bornat \emph{et al.} gave an early attempt to reason about graph algorithms in separation logic in a more general way~\cite{bornat:aliasing04}.

Reynolds was the first to document the overlapping conjunction~$**$, although he did not present any strategy to reason about it using Hoare rules~\cite{rey-slnotes}.  Gardner \emph{et al.} were the first to reason about a program using $**$ in Javascript~\cite{GardnerMS12}.  Raad \emph{et al.} used $**$ to reason about a concurrent spanning algorithm using a kind of ``concurrent localization''~\cite{RaadVG15}.  Sergey \emph{et al.} also verified a concurrent spanning tree algorithm and mechanized their proofs in Coq~\cite{ilya-graphs}.

Almost a decade after Yang verified Schorr-Waite on paper, Dafny automated its verification~\cite{Leino10}.

%\paragraph{Alternative fixpoint constructions.}
%Appel and McAllester defined an alternative ``contractive'' fixpoint that is sometimes used to define recursive predicates in separation
%logic~\cite{appel:fixpoint}.  In Appendix~\ref{apx:appelfixpiont} we explain why Appel and McAllester's  is also unsuitable to define graphs.

\paragraph{Verification tools.}
Our work heavily interacts with the Floyd~\cite{appel:programlogics} and HIP/SLEEK \cite{chin:hipsleek} verification tools.  Like Floyd, Charge! uses Coq tactics to work with a shallow embedding of higher order separation logic, but focuses on OO programs written in Java/C\#~\cite{bengtson:charge}.  A  more automated approach to verification of low level programs using Coq is implemented in the Bedrock framework \cite{chlipala:bedrock}.

Many automated verification tools also use separation logic in a forward reasoning style as does HIP/SLEEK, including Smallfoot~\cite{berdine:smallfoot}, jStar~\cite{DistefanoP08}, and Verifast~\cite{jacobs:verifast}.  One of HIP/SLEEK's distinguishing features is good support for user-defined inductive predicates rather than a library of pre-defined predicates for lists, trees etc.

Dafny~\cite{Leino10} and KeY~\cite{Beckert:2007} are two notable verifiers not based on separation logic; KeY uses an interactive verifier while Dafny pursues more automation with Z3~\cite{Moura2008}.

\paragraph{Mechanized mathematical graph theory.} There is a long history,
going back at least 25 years, of mechanized reasoning about mathematical
graphs~\cite{wong1991}.  The most famous mechanically verified ``graph theorem''
is the Four Color Theorem~\cite{gonthier2005computer}; however
the development actually uses hypermaps instead of graphs.
Noschinski built a graph library in Isabelle/HOL whose formalization
is the closest to ours~\cite{Noschinski2015}, \emph{e.g.} supporting
graphs with labeled and parallel arcs.

Noschinski and Dubois \emph{et al.} used proof assistants to design verifiable
checkers for solutions to graph problems~\cite{noschinski2015formalizing,dubois2015graphes}.
Yamamoto \emph{et al.} and Bauer and Nipkow use an inductive
encoding of graphs to formalize planar graph theory~\cite{yamamoto1995formalization,bauer20025}.


%{\color{magenta}
%5.2. An alternative way of verifying marking program is reasoning about the whole history of marking operations. The disadvantage of it is that it currently needs more work in a Hoare logic framework. The advantage of it is that its reasoning structure are more similar with the way we understand it in our first algorithm class.
%5.3. I take some effort on garbage collector like graph structural. Though it is only connecting this special structural with 5.1.1 and 1.3, it takes much time and it is not finished yet.}


