In the future we plan to improve the pure reasoning of graphs and
similar data structures, with a particular focus on automation.  We have
also begun to investigate integrating our techniques into the 
HIP/SLEEK toolchain~\cite{chin:hipsleek}, which, as compared to VST, 
provides more automation at the cost of lower expressivity.  We
are also interested in investigating better ways to handle the
kinds of undefined behavior upon which real~C systems code sometimes
relies.

\hide{\color{magenta}We are in the process of verifying a garbage
collector for the ``CertiCoq'' project, which is building
a certified compiler from Gallina to Clight. We would like to investigate
using our externally verified lemmas in HIP/SLEEK to verify code such as fast
exponentiation and more graph algorithms. We also would like to make
the interface between Coq and H/S simpler and cleaner.
One final direction we would like to investigate is using our new
connection to Coq to have H/S output certificates as it
verifies programs so that the system becomes more trustworthy.}

Our main contributions were as follows.  We developed a mathematical
graph library that was powerful enough to reason about graph-manipulating
algorithms written in real~C code.  We connected these mathematical graphs
to spatial graphs in the heap via separation logic.  We developed 
localization blocks to smoothly reason about a local action's effect on
a global context in a mechanized context, including a robust treatment
of modified program variables and existential quantifiers in postconditions.
We demonstrated our techniques on several nontrivial examples, including union-find
and spanning tree.  Our flagship example was the verification of the garbage collector 
for the CertiCoq project, during which we found two places in which the~C semantics
is too weak to define an OCaml-style GC.  We integrated our techniques into the
VST toolset. % and reported on our proof development.

