Over the last fifteen years, separation logic has facilitated great strides
in verifying programs that manipulate tree-shaped data structures.
\citep{berdine:smallfoot,chin:hipsleek,jacobs:verifast,
chlipala:bedrock,bengtson:charge,appel:programlogics}.
Unfortunately, programs that manipulate graph-shaped data structures
(\emph{i.e.} structures with \emph{intrinsic sharing}) have proved harder to verify.
Indeed, such programs were formidable enough that many of the early
landmark results in separation logic devoted substantial effort to verifying
single examples such as Schorr-Waite~\cite{hongseok:phd} or
union-find~\cite{neelthesis} with pen and paper.
More recent landmarks have moved to a machine-checked context, but have still
been devoted to either single examples or to classes of closely-related examples
such as garbage collectors~\cite{gcexample3,cakemlgc}.
These kinds of examples tend to require a large number
of custom predicates and subtle reasoning, which generally
does not carry over
to the verification of other graph-manipulating programs.

In contrast, we present a toolkit for verifying graph-manipulating programs in a
machine-checked context. Our techniques are \emph{general} in that they handle a diverse
range of programs, and \emph{modular} in that they encourage code
reuse (\emph{e.g.} facts about reachability) and separation of concerns
(\emph{e.g.} between abstract mathematical graphs and concrete representations
in the heap).
Our techniques are \emph{powerful} enough to reason about real C code as compiled by
CompCert~\cite{leroy:compcert}, and also
\emph{lightweight} enough to integrate into the Verified
Software Toolchain (VST)~\cite{appel:programlogics} without major reengineering.
CompCert and VST are distributed via opam and the CoqIDE installer, so they have a sizable userbase that can take advantage of our techniques. %``out of the box''.
Finally, our techniques \emph{scale} well beyond short toy programs: we certify the
correctness of a generational garbage collector for the CertiCoq
project~\cite{certicoqpaper} ($\approx400$ rather devilish lines of C).

We proceed in three steps. First, we develop a mathematical graph library that is general enough to reason about a wide variety of algorithms and expressive enough to describe the behavior of these algorithms in real machines.  We modularize this library carefully so that common ideas---\emph{e.g.}~subgraphs, reachability, and isomorphism---can be efficiently reused
in different algorithms.  Second, we use separation logic to express how these abstract graphs
are actualized in the heap as concrete graphs in a way that facilitates the reuse of key definitions and theorems across algorithms.  Third, we develop a notion of \emph{localization blocks} that
enables modular reasoning in our Hoare proofs, even in the presence of the
implicit sharing intrinsic to graphs, by using our \textsc{Localize} rule:

% old two-column version:
\hide{
\begin{equation}
\label{eq:localize}
\begin{array}{@{}l@{}}
\infrule{Localize}
{G_1 |- L_1 * R \\
\{ L_1 \} ~ c ~ \{ \exists x.~ L_2 \} \\
R |- \forall x.~ (L_2 --* G_2) }
{\{ G_1 \} ~ c ~ \{ \exists x.~ G_2 \}} {(\dagger)} \\
[3pt]
(\dagger)~ \mathit{freevars}(R) \cap \MV(c) = \emptyset
\end{array}
\end{equation} \marginpar{Can we typeset this a little better?}
} % end hide

% new one-column version:
\vspace{-1em}
\begin{equation}
\label{eq:localize}
\inferrule[Localize]	
{G_1 |- L_1 * R \\
\{ L_1 \} ~ c ~ \{ \exists x.~ L_2 \} \\
R |- \forall x.~ (L_2 --* G_2) }
{\{ G_1 \} ~ c ~ \{ \exists x.~ G_2 \}} \; \FV(R) \cap \MV(c) = \emptyset \qquad
\vspace{-0.3em}
\end{equation}
%\marginpar{\tiny \color{blue} Maybe this should be moved to \S2? Just formula for intro?}
\textsc{Localize} connects the ``local'' effect of a command~$c$, \emph{i.e.}
transforming~$L_1$ to~$L_2$, with its ``global'' effect, \emph{i.e.} from~$G_1$ to~$G_2$.
\textsc{Localize} is a more general version of the well-known \textsc{Frame} rule,
which does the same task in the simpler case when $G_i = L_i * F$
for some frame~$F$ that is untouched by~$c$.
\textsc{Localize} can handle the more general transformation if we can find a \emph{ramification frame}~$R$ that satisfies a pair of
delicately-stated entailments\footnote{Readers less familiar with the separating implication $P --* Q$, also known as \emph{magic wand}, can refer to its semantics in Figure~\ref{fig:seplogsem} (page~\pageref{fig:seplogsem}), which also models the other separation logic operators
we use in this paper.  The key proof rule for magic wand is its adjointness
with the separating conjunction:~$(P * Q |- R) \Leftrightarrow (P |- Q --* R)$.} and
the side condition on modified local program variables.
\textsc{Localize} %Localization blocks upgrade
upgrades the \textsc{Ramify} rule~\cite{hobor:ramification} in two key ways:
support for existential
quantifiers in postconditions and smoother treatment of modified program variables.

%as a kind of proof pattern or framework to verify graph-manipulating programs on pen and paper~\cite{hobor:ramification}.  The major focus of this paper is to develop methods to verify realistic graph programs in a mechanized context.  We do so by upgrading the theory of ramification and by developing a general and modular library for graph-related reasoning in separation logic.  We incorporate our approach into two sizeable separation logic-based verification tools: the Floyd system of the Verified Software Toolchain (VST)~\cite{appel:programlogics} and the HIP/SLEEK program verifier~\cite{chin:hipsleek}.  VST and HIP/SLEEK inhabit quite different points in the design space for verification tools, with VST primarily focusing on heavily human-guided verifications with an emphasis on end-to-end machine-checked proofs, and HIP/SLEEK focusing on more automation.  Despite these differences, the vast majority of our Coq code base is shared between them,
%giving us hope that our work will be applicable to other tools.

%% \marginpar{\color{magenta} computable mathgraphs, null, pregraphs Problem with ``later'' not being precise.}
Our contributions are organized as follows:
\begin{itemize}
\item[\S\ref{sec:orientation}] We use the classic ``union-find'' disjoint set algorithm to show how our three key ingredients---mathematical graphs, spatial graphs, and localization blocks---come together to verify
graph-manipulating algorithms.
%\marginpar{\tiny \color{blue} TODO: check.}
%Although we use it primarily as a running example,
To the best of our knowledge this is the first machine-checked verification of this algorithm that starts with real C code.
%We introduce \emph{localization blocks} as a notation for \textsc{Localize} in 
%decorated programs. 
We briefly review the seven programs we have verified to give a sense 
of the breadth of algorithms our system can tackle.
\item[\S\ref{sec:localizations}] We show that \textsc{Localize} and \textsc{Frame} are co-derivable.  We illustrate a delicate technique to properly handle modified 
local variables. We show a mark-graph
program that explores a graph in a fold/unfold style, and discuss the utility of linked
existentials in postconditions.  

%{\color{magenta}Existentials for fun and profit: we upgrade Hobor and Villard's \infrulestyle{Ramify} rule to handle both modified program variables and existential quantifiers in postconditions more gracefully.
%    We briefly discuss four additional VST-certified examples to give a sense of the breadth of algorithms we can verify: marking a DAG, marking a graph, an array-based version of union-find, and pruning a graph into a spanning tree.} %, and making a structure-preserving copy of a graph.
\item[\S\ref{sec:mathgraph}] We develop a general and modular framework of mathematical graphs powerful enough to support realistic verification in a mechanized context.  We give a sampling of key definitions.
\item[\S\ref{sec:spacegraph}] We suggest that the Knaster-Tarski fixpoint~\cite{tarski:fixpoint} cannot define a usable separation logic graph predicate.  We propose a better definition for general spatial graphs that still enjoys a ``recursive'' fold/unfold.  We prove general theorems about spatial graphs that are 
organized in a modular way that can be utilized in multiple flavors of separation logic. %{\color{magenta} Line about modularity?}
%\item[\S\ref{sec:hipsleek}] We explain how we modified HIP/SLEEK to introduce ramifications when programs modify data structures with intrinsic sharing and to automatically discharge the associated obligations using Coq-verified external lemmas.
%\item[\S\ref{sec:vst}] We explain how we integrated ramification into VST by developing .
\item[\S\ref{sec:certigc}] We discuss our flagship example, the certification of the CertiCoq garbage collector (GC). We identify two places where the semantics of C is too weak to define an OCaml-style GC. We also find and fix a rather subtle overflow error in the original C code for the GC, thereby justifying the effort of developing machine-checked proofs of correctness.
%\marginpar{\tiny \color{blue} Bit funny to put these two in the same section.}
\item[\S\ref{sec:development}] We discuss how our techniques are integrated into the
``Floyd'' module of VST, a separation-logic based engine to help users verify
CompCert~C programs, via two new Floyd tactics \li{localize} and \li{unlocalize}.
We also document statistics related to our overall development.
\item[\S\ref{sec:related}] We discuss related work.
\item[\S\ref{sec:conclusion}] We discuss directions for future work and conclude.
\end{itemize}
All of our results are machine checked in Coq and are available as an artifact 
\href{https://zenodo.org/record/3368779}{online}.
An extended version of this paper featuring three appendices is also available 
online~\cite{onlinepaper}.
